\cleardoublepage{}
\phantomsection{}
\addcontentsline{toc}{chapter}{Acknowledgements}
\begin{acknowledgement}
  Before diving into the main content of this thesis, I would like to thank all the people who have made this work flourish and become what it is today.

  First and foremost, I would like to thank Franck and Julien for supervising this thesis, and for guiding me as they have done over the past three and a half years. Our theoretical and technical discussions, as well as our numerous drawings on whiteboards during our weekly meetings have been the main contributing factors of this thesis.

  Secondly, I would like to thank Guillermo Gallego and Vincent Lepetit for reviewing this thesis, and for sending me their precious comments to improve it. I would also like to thank Rémi Boutteau, Véronique Cherfaoui, and Jean Martinet for examining this work, as well as Pascal Vasseur and Philippe Xu for providing an external view on my work across these three years.

  Thirdly, I would like to thank all the fellow PhD students at Heudiasyc I have had the privilege of working with during all my time at the lab: Maxime \& Maxime, Ali, Antoine, Matthieu, Rémy, Loïc, and Michaël. I would also like to thank Thierry and Stéphane, for their precious technical assistance and availability. More generally, I would like to thank all the members and personnel of Heudiasyc, SIVALab, and UTC, whom it has always been a pleasure to work with.

  Finally, I would like to thank from the bottom of my heart my parents and my brother Benoît, for constantly giving me precious advices, strength, and emotional support over all these years.
\end{acknowledgement}
